---
title: "Beta Binomial Model - Your Use Case"
format: html
editor: visual
---

# Welcome to Today's Lecture

```{r}
library(tidyverse)
library(bayesrules)
```

We have done a deep-dive into the [Beta Binomial model for Michellele's election support](https://www.bayesrulesbook.com/chapter-3) with the shiny app in the `beta_binomial_election_app.R` script. Now, I would like you and a teammate to create a use-case that is appropriate for the usage of this model.

## Example

This example will help guide you in your use case, I will leave a skeleton outline of the bottom of this qmd for you to fill out.

### Prior

Every four years, Americans go to the polls to cast their vote for President of the United States. Consider the following scenario. “Michelle” has decided to run for president and you’re her campaign manager for the state of Minnesota. As such, you’ve conducted 30 different polls throughout the election season. Though Michelle’s support has hovered around 45%, she polled at around 35% in the dreariest days and around 55% in the best days on the campaign trail

π = Michelle's true election support

```{r}
plot_beta(45, 55)
```

Statement = "Before any new poll, we believe there's a 95% chance Michelle's true support π is between 35% and 55%, with 45% as our single best guess."

Prior effective sample size = 100 pseudo voters

### Likelihood

**New data arrives:** We commission a new poll of *n* Minnesotans and record *y* , the number who support Michelle. We poll 50 voters in which 30 (60%) supported her -- noticeably higher than her 45% prior average.

```{r}
plot_binomial_likelihood(y = 30, n = 50)
```

Likelihood Question: "For each possible value of π — Michelle's true underlying support — how compatible is that value with having produced exactly 30 out of 50 in this poll?"

Likelihood Statement: "Taken on its own, this poll makes values of π near 60% far more plausible than values near 45% — the data alone points to a higher number than we'd been expecting"

### Posterior

```{r}
plot_beta_binomial(alpha = 45, beta = 55, y = 30, n = 50)
```

Statement = "There's a 95% probability that Michelle's true support π lies between 42.0% and 58.0%."

Weight = prior (worth \~100 pseudo-voters) contributes 67% of the weight, and our new poll (n = 50 voters) contributes the other 33%.

## Your Turn

With your teammate, invent a use-case appropriate for the Beta Binomial Model. You do not need to find a dataset, just follow the kind of thinking from the text and example. Repeat the steps above describing in context your prior, likelihood, and posterior with statements and plots. Tips:

- Come up with 3 -5 ideas on your own

- Sketch them out and think critically in how they differ from the example but may still fit the Beta Binomial Modeling architecture

- Pitch your sketches to a free AI Chat of your choosing (I use Claude) and have it critique your use case to see if the Beta Binomial Model is appropriate and accurately addresses your question

- Note the assumptions/limitations of your model based on AI feedback

- **Remember**: "[All models are wrong, but some are useful](https://en.wikipedia.org/wiki/All_models_are_wrong)"

Things to keep on mind:

- Is π a single, clearly defined quantity — not "kind of about deposits, kind of about attendance"?

- Does your data collection method actually measure that population, or a biased subset of it (who responds, who's easy to ask)?

- Is π genuinely uncertain — or is it something you could already know exactly (a completed census), making the model unnecessary?

- Is your prior based on real past counts (pseudo-counts) or an elicited belief — and can you say why you chose that one?

## AI Prompt:

Can you critique this idea as a beta binomial use case: A Teacher is deciding which operating system to make the examples with the other operating systems will have linked guides on the page. The class last year was 12 apple users and 43 windows users The class this year was 22 apple users and 47 windows users Which operating system should next year's class be primarily geared towards.

## AI Response:

Solid instinct, but there's a real trend issue baked into your data that a naive beta-binomial pooling would hide.

**What you'd do:** Treat "P(Apple)" as your parameter, put a prior (maybe Beta(1,1) or informed by industry stats), update with year 1 (12/55), update again with year 2 (22/69), get a posterior Beta(35, 96), mean ≈ 26.7%. Decide Windows-primary since posterior mass is overwhelmingly below 50%.

**Where it's shakier:**

1.  **Stationarity assumption.** Beta-binomial pooling implicitly assumes both years are draws from the *same* underlying p. But your data shows real drift: 21.8% → 31.9% Apple share. That's a 10-point jump, not noise you'd want to average away. Pooling treats last year and this year as equally informative about *next* year, which throws away the trend.

    **Fix options:**

    - Weight recent data more (e.g., use year 2 as your effective prior mass, or apply a discount factor to year 1 counts before pooling).

    - Go hierarchical: model each year's count as Binomial(n_t, p_t) with p_t drawn from a shared distribution, which lets you see between-year variance instead of masking it.

    - Simplest defensible tweak: just use year 2 alone (or year 2 with a weak prior) as your best estimate of "current" p, and mention year 1 only as evidence of an upward trend in Apple share, i.e., flag that even 32% may undershoot next year.

    **The decision itself isn't just a point estimate.** "Which OS to primarily gear toward" is a decision problem, not just an inference problem — you're implicitly minimizing something like "cost of a student needing the linked guide instead of the primary walkthrough." Beta-binomial gives you the posterior on p; you still need to state the loss function (majority-rules threshold at 50% is one choice, but if Apple's share is trending up, you might want a buffer, like only committing to Windows-primary if P(p_Apple \< 0.4) is high).

    **Binary OS assumption.** You've excluded Linux/other entirely — fine if genuinely negligible, but worth stating as an assumption.

Bottom line: beta-binomial is a fine tool for the "what's my best estimate of p" piece, but pooling two years naively will understate Apple's momentum. If the trend continues, Windows may still be safely primary this year, but you're closer to the line than the pooled 26.7% estimate suggests — I'd flag that explicitly rather than let the model average it away.

### Prior

Set the scene...

A Teacher is deciding which operating system to make the examples with the other operating systems will have linked guides on the page.

The class last year was 12 apple users and 43 windows users

The class this year was 22 apple users and 47 windows users

Which operating system should next year's class be primarily geared towards.

define π

```{r}
plot_beta(alpha = 12, beta = 43)
  
```

Statement = "Before any new data we believe there's a 95% chance π is between 12% and 34%, with 22% as our single best guess."

Prior effective sample size = 55

### Likelihood

**New data arrives:** We commission new data of *n* and record *y* , the number .... We gather x data points in which y (x%) ... -- noticeably higher/lower than the x% prior average.

```{r}
plot_binomial_likelihood(y = 22 , n = 69 )
```

Likelihood Question: "For each possible value of π —.... — how compatible is that value with having produced exactly y out of n in this data collection?"

Likelihood Statement: "Taken on its own, this data makes values of π near 34% far more plausible than values near 12% — the data alone points to a higher number than we'd been expecting"

### Posterior

```{r}
plot_beta_binomial(alpha = 12 , beta = 43 , y = 22 , n = 69)
```

Statement = "There's a 95% probability that π lies between 20% and 35.6%."

Weight = prior (worth \~55 students) contributes 44% of the weight, and our new data (n =69 ) contributes the other 56%.
